---
title: "Naive intervals after a recorded, effect-blind search over emulated protocols"
subtitle: "A simulation study answering the recorded-search component of catalog problem PRO-02"
author: "Ahmad Sofi-Mahmudi"
date: 2026-09-23
bibliography: references.bib
csl: ../../../vancouver.csl
link-citations: true
number-sections: true
execute:
  echo: false
  warning: false
  message: false
format:
  gfm:
    toc: false
  pdf:
    documentclass: article
    geometry: margin=1in
    fig-pos: "H"
  odt: default
---

```{r setup}
R <- "../results/replication"
f <- function(x, d = 3) formatC(x, format = "f", digits = d)
gd <- read.csv(file.path(R, "global-deficits.csv"))
cal <- read.csv(file.path(R, "calibration.csv"))
perf <- read.csv(file.path(R, "performance.csv"))
rem <- read.csv(file.path(R, "remedy-performance.csv"))
els <- read.csv(file.path(R, "effective-library-size.csv"))
mvr <- read.csv(file.path(R, "multiverse-risk.csv"))
auc <- read.csv(file.path(R, "multiverse-auc.csv"))
wid <- read.csv(file.path(R, "width-comparisons.csv"))
first <- read.csv("../results/global-deficits.csv")
first_cal <- read.csv("../results/calibration.csv")
g <- function(e, col) gd[[col]][gd$estimand == e]
nv <- perf[perf$method == "naive", ]
blind <- nv[nv$selector %in% c("workability", "prognostic_high", "prognostic_low", "outcome_blinded"), ]
seek <- nv[nv$selector == "effect_seeking", ]
cov_by <- aggregate(coverage ~ selector, nv, mean)
rk <- function(cat) range(mvr$noncoverage_risk[mvr$category == cat])
rm <- function(p, col) rem[[col]][rem$policy == p]
```

# Abstract {.unnumbered}

**Background.** Emulation protocols are often revised after investigators see
the data, and the reported interval ignores that search. Catalog entry PRO-02
states that this invalidates coverage. We measured how much coverage a naive
interval loses when the search is recorded and the selector never sees the
treatment effect.

**Methods.** Thirty-six generated-data cells crossed cohort size, expected event
count, library size (4 or 16 candidate protocols over age thresholds and outcome
horizons) and treatment effect, with 2,000 replicates each. Effect-blind
selectors chose a protocol by arm sizes and pooled events, by prognostic
associations, or by covariate balance; an effect-seeking selector was the
positive control. Estimands were the global coverage deficits D_all and
D_nonboundary against a registered materiality threshold of 0.02. This is a
registered replication: the first run failed its calibration gate by chance.

**Results.** All 360 fixed-candidate evaluations passed the calibration gate.
D_all was `r f(g("D_all", "deficit"), 4)` (simultaneous 95% interval
`r f(g("D_all", "lower"), 4)` to `r f(g("D_all", "upper"), 4)`) and
D_nonboundary `r f(g("D_nonboundary", "deficit"), 4)`
(`r f(g("D_nonboundary", "lower"), 4)` to `r f(g("D_nonboundary", "upper"), 4)`).
The effective library size was at most `r f(max(els$effective_library_size), 2)`.
Noncoverage rose to `r f(max(rk("above_2")))` when candidate estimates spanned
more than two standard errors. The effect-seeking control lowered coverage to
`r f(min(seek$coverage))` in the worst cell.

**Conclusions.** In this design a recorded, effect-blind search costs naive
intervals almost nothing on average, because the candidates barely differ. When
they do differ, the selected interval undercovers.

# Introduction

Target trial emulations specify eligibility, treatment strategies, follow-up and
outcomes before analysis, and reporting guidance asks that the specification and
its revisions be recorded [@cashin2025target]. In practice thresholds and
horizons are revised after inspecting the data. Even a selector that never sees
the effect estimate can use outcome frequency or covariate-outcome associations,
and those are correlated with the estimate. The interval reported for the chosen
protocol conditions on a choice that was itself data dependent.

PRO-02 treats this as a coverage problem. This study measures its size for the
part that can be simulated: a finite, recorded search whose selector is blind to
the treatment effect. Undocumented, sequential development by an analyst who has
seen results is outside its scope.

# Methods

The study follows the ADEMP structure [@morris2019simulation]. The design was
reviewed and revised before any code was written.

## Data-generating mechanism

Age is 55 + 10Z with Z a standard normal truncated at ±2; S and C are binary
covariates and B continuous. Baseline initiation depends on Z, S, C and B, and
time to event follows a constant-hazard model with the same covariates. The
treatment effect is null, a hazard ratio of 0.75, or a hazard ratio
exp(−0.288 + 0.25Z) that changes with age, so candidate protocols with different
age thresholds target different effects. The baseline hazard was solved, before
the study and from event counts only, so that the anchor protocol (age at least
55, 36 months) has 35, 70 or 140 expected pooled events. There is no loss to
follow-up. The four-candidate library crosses age thresholds of 50 and 55 with
horizons of 36 and 60 months; the sixteen-candidate library crosses thresholds
of 45 to 60 with horizons of 24 to 60 months.

## Estimands and selectors

For each candidate the estimand is the risk difference at its horizon among
people meeting its age threshold. Each protocol's effect was estimated by Hajek
inverse probability weighting with a correctly specified propensity model and a
Wald interval. Selectors: a workability rule (the first candidate with at least
100 per arm and 70 pooled events); two prognostic-association rules, whose fixed
score tracks the outcome more or less closely; an outcome-blinded rule that
picks the eligibility threshold with the best covariate balance; and, as a
positive control, the candidate with the most beneficial standardized estimate.
Two remedies were evaluated: Bonferroni over the declared library and a
fifty-fifty split between a design half that selects and an analysis half that
estimates.

## Decision rule

A calibration gate comes first: every declared candidate in every cell must
have fixed-candidate coverage compatible with [0.93, 0.97], judged by a Wilson
interval at familywise confidence 1 − 0.05/360, and failure probability
compatible with at most 0.05. If it passes, the problem is materially real if
both simultaneous 95% lower limits of D_all and D_nonboundary exceed 0.02, not
materially real if both upper limits are below 0.02, and mixed otherwise.

## Why this is a replication

The first run used the same design with a gate that required all 360 coverage
estimates to lie inside [0.93, 0.97]. One came in at
`r f(min(first_cal$coverage), 4)`, 0.26 Monte Carlo standard errors outside, and
the registered classification was held at uninformative. A gate applied without
allowance for Monte Carlo error to 360 estimates fails almost surely even when
every candidate is calibrated. The replication protocol, committed before any
of its replicates were drawn, changed only the seed and the gate. The first
run's deficits, `r f(first$deficit[first$estimand == "D_all"], 4)` and
`r f(first$deficit[first$estimand == "D_nonboundary"], 4)`, were also far below
the threshold.

# Results

## Calibration and the primary comparison

All `r nrow(cal)` fixed-candidate evaluations passed the gate; fixed-candidate
coverage ran from `r f(min(cal$coverage), 4)` to `r f(max(cal$coverage), 4)`
and no candidate failed to estimate. D_all was `r f(g("D_all", "deficit"), 4)`
(`r f(g("D_all", "lower"), 4)` to `r f(g("D_all", "upper"), 4)`) and
D_nonboundary `r f(g("D_nonboundary", "deficit"), 4)`
(`r f(g("D_nonboundary", "lower"), 4)` to
`r f(g("D_nonboundary", "upper"), 4)`). Both upper limits lie more than 0.015
below the threshold. The registered branch is that the recorded-search problem
is not materially real in this design. The nonboundary deficit is detectably
above zero but equals about a fifth of a percentage point.

## Why the average deficit is small

Nested age thresholds and horizons produce highly correlated candidates. The
effective library size (the squared sum of the eigenvalues of the candidates'
correlation matrix divided by the sum of their squares) was
`r f(min(els$effective_library_size[els$K == 4]), 2)` to
`r f(max(els$effective_library_size[els$K == 4]), 2)` with 4 candidates and
`r f(min(els$effective_library_size[els$K == 16]), 2)` to
`r f(max(els$effective_library_size[els$K == 16]), 2)` with 16. Choosing among
them is close to not choosing.

## The design can detect a loss

The effect-seeking selector lowered naive coverage to
`r f(mean(seek$coverage), 3)` on average and to `r f(min(seek$coverage))` to
`r f(max(seek$coverage[seek$K == 16]))` across the 16-candidate cells, against
`r f(min(cov_by$coverage[cov_by$selector != "effect_seeking"]))` to
`r f(max(cov_by$coverage[cov_by$selector != "effect_seeking"]))` for the
effect-blind selectors (@tbl-coverage).

```{r}
#| label: tbl-coverage
#| tbl-cap: "Mean naive coverage of the selected protocol's interval, by selector and treatment effect."
w <- aggregate(coverage ~ selector + effect, nv, mean)
w <- reshape(w, idvar = "selector", timevar = "effect", direction = "wide")
names(w) <- sub("coverage.", "", names(w), fixed = TRUE)
knitr::kable(w, digits = 3, row.names = FALSE)
```

## The average hides a conditional failure

Naive noncoverage depended on how far the candidates' estimates spread
(@tbl-multiverse). When the range across candidates was below one median
standard error, noncoverage was `r f(min(rk("below_1")))` to
`r f(max(rk("below_1")))`; between one and two, `r f(min(rk("1_through_2")))`
to `r f(max(rk("1_through_2")))`; above two, `r f(min(rk("above_2")))` to
`r f(max(rk("above_2")))`. The standardized range predicted noncoverage with an
AUC of `r f(min(auc$auc), 2)` to `r f(max(auc$auc), 2)`.

```{r}
#| label: tbl-multiverse
#| tbl-cap: "Naive noncoverage by the standardized range of the candidate estimates (range divided by the median candidate standard error)."
m <- reshape(mvr[, c("selector", "category", "noncoverage_risk")],
             idvar = "selector", timevar = "category", direction = "wide")
names(m) <- c("selector", "1 to 2", "above 2", "below 1")
knitr::kable(m[, c("selector", "below 1", "1 to 2", "above 2")], digits = 3,
             row.names = FALSE)
```

## Remedies

Bonferroni intervals for the selected protocol covered
`r f(rm("selected_bonferroni", "coverage"), 4)`, above nominal. Sample splitting
covered `r f(rm("sample_split", "coverage"), 4)`, with intervals a median
`r f(median(wid$median_within_candidate_width_ratio), 2)` times as wide as the
full-sample interval for the same candidate, close to the square root of two.

# Discussion

A recorded search whose selector cannot see the effect does little damage to
average coverage when the candidates are variations on one protocol. The damage
concentrates in the replicates where the candidates disagree, which is also
where the choice matters. Reporting every candidate's estimate, as a multiverse
display, therefore carries information about the reliability of the selected
interval that the interval itself does not.

Bonferroni and sample splitting both restore coverage. Bonferroni overshoots;
splitting pays in width about what halving the data predicts.

## What this study does not establish

The result covers deterministic recorded searches over nested age thresholds and
horizons, libraries of 4 or 16 highly correlated candidates, effect-blind
selectors and correctly specified weights. It does not cover sequential protocol
development by an analyst who has seen results, other protocol components such
as grace periods or treatment definitions, libraries of weakly correlated
protocols, or misspecified weights. The replication's gate controls false
failure across 360 comparisons and would pass a candidate miscalibrated by less
than about two percentage points.

## Review

The design was reviewed and revised before any code was written. The first run
was reviewed twice; the reviewer held its classification at uninformative and
recommended a separately registered replication, which is this study. The
replication's results have not yet been reviewed.

# Data and code availability {.unnumbered}

Code, both protocols and result files: <https://github.com/choxos/TTE-open-problems/tree/main/studies/PRO-02-ben-02-recorded-protocol-search>.

# References {.unnumbered}
