---
title: "Unrecorded eligibility changes who is eligible more than it changes the effect among them"
subtitle: "A simulation study answering the structural part of catalog problem ELG-01"
author: "Ahmad Sofi-Mahmudi"
date: 2026-09-24
bibliography: references.bib
csl: ../../../vancouver.csl
link-citations: true
number-sections: true
execute:
  echo: false
  warning: false
  message: false
format:
  gfm:
    toc: false
  pdf:
    documentclass: article
    geometry: margin=1in
    pdf-engine: xelatex
    mainfont: "STIX Two Text"
    fig-pos: "H"
  odt: default
---

```{r setup}
R <- "../results"
f <- function(x, d = 3) formatC(x, format = "f", digits = d)
dec <- read.csv(file.path(R, "decision.csv"))
perf <- read.csv(file.path(R, "performance.csv"))
prev <- read.csv(file.path(R, "prevalence-performance.csv"))
td <- read.csv(file.path(R, "target-discrepancy.csv"))
perf$mv <- paste(perf$method, perf$variant)
prev$mv <- paste(prev$method, prev$variant)
full <- perf[!grepl("_m50$", perf$method), ]
pair <- prev[prev$effect == "heterogeneous" & prev$kappa == 1.1 &
             prev$validation_n == 100 & prev$m == 0.5, ]
pe <- tapply(pair$truth_pe, pair$lambda_key, mean)
mar <- pair[pair$mv %in% c("mar_ipaw primary", "mi_continuous m20",
                           "mi_indicator m20") & pair$lambda_key == "lambda_log4", ]
val <- pair[pair$information_set == "validation", ]
het <- td[td$effect == "heterogeneous" & td$m == 0.5 & td$kappa == 1.1 &
          td$validation_n == 100, ]
```

# Abstract {.unnumbered}

**Background.** When the measurement that defines eligibility is missing for
some patients, an emulation cannot tell from its records who was eligible.
Catalog entry ELG-01 states that latent eligibility makes the target population
unidentified without validation data. We asked whether that ambiguity changes
the treatment effect materially, and how estimators with and without internal
validation perform.

**Methods.** Paired full-data laws with identical observed records but
different eligibility among people with unrecorded measurements (a shift of
log odds λ of 0, log 2 or log 4) were crossed with the share unrecorded (0.20,
0.50), the expected validation sample (100, 400), a homogeneous or
heterogeneous treatment effect, and whether comorbidity, the effect modifier,
enters eligibility directly: 48 scenarios, 4000 replicates each, 4000 people
per replicate. The registered structural contrast Δ~ID~ compares the true
risk difference among the eligible at λ = log 4 and λ = 0 in the pair most
likely to show a difference.

**Results.** Δ~ID~ was `r f(dec$delta_id, 5)` (90% interval
`r f(dec$ci90_lo, 6)` to `r f(dec$ci90_hi, 6)`), inside the registered
equivalence band of ±0.005: not materially different. The same shift moved the
true prevalence of eligibility from `r f(pe[["lambda_0"]])` to
`r f(pe[["lambda_log4"]])`, which estimators assuming eligibility is missing at
random missed by up to `r f(max(abs(mar$bias)))`. Every estimator's risk
difference was unbiased within ±0.005 in every scenario.

**Conclusions.** In this design, unrecorded eligibility changes how many people
are eligible far more than the effect among them. Validation data are needed to
count the eligible; the effect estimate is robust unless the hidden eligibility
shifts the share of an effect modifier substantially.

# Introduction

Eligibility criteria defined by a laboratory value or a clinical assessment are
often unrecorded for part of the cohort. Restricting to recorded eligibles, or
imputing eligibility under a missing-at-random assumption, then targets a
population that may differ from the truly eligible one
[@tompsett2023missing; @benz2025missing]. Catalog entry ELG-01 states that
without validation data the eligible population is not identified. That is true
by construction: two data-generating laws can produce identical records and
different eligibility among the unrecorded. Whether the difference matters for
the effect estimate is a separate question, and it is the one this study asks.

# Methods

The study follows the ADEMP structure [@morris2019simulation]; the protocol and
decision rule were registered before any replicate was drawn.

## Data-generating mechanism

Each replicate has 4000 people with baseline covariates age, sex, specialist
access H, comorbidity C and recorded treatment intent, and a single baseline
treatment decision. A clinical recording indicator R is drawn with P(R = 0)
calibrated to m (0.20 or 0.50). A latent laboratory value G = 45 + 5T defines
eligibility E = 1 when G is at least 45, where T is logistic with linear predictor
0.85 − 0.65Z − κC + 0.25S + 0.50H + λ(1 − R); κ is 0 or 1.10. Changing λ
changes eligibility only among people whose value is unrecorded, and leaves
every observed variable bitwise identical, so paired λ arms are observationally
equivalent. The 60-month outcome has baseline risk depending on covariates and
a treatment effect of −0.06 for everyone (homogeneous) or −0.04 without and
−0.12 with comorbidity (heterogeneous). A Bernoulli validation sample with known
inclusion probability reveals G for some unrecorded people.

## Estimands and methods

The primary estimand is the 60-month risk difference among the truly eligible,
RD~E~; the complete-case analysis is evaluated against the risk difference
among recorded eligibles, and the difference between the two is reported as a
target discrepancy. Truth was computed from ten million simulated people per
law. Methods without validation: complete case; multiple imputation of G or of
E under missing at random (M = 20, with an M = 50 check); inverse probability
of ascertainment weighting; and a pattern-mixture sensitivity analysis at
assumed λ\* of 0, log 2 and log 4. Methods with validation: known-design
two-phase weighting, validation-anchored fractional imputation and an augmented
two-phase estimator. An oracle with full eligibility is the benchmark.

## Decision rule

The structural result uses one prespecified pair: heterogeneous effect,
m = 0.50, κ = 1.10, λ = log 4 against λ = 0. It is material if the 95% paired
Monte Carlo interval for Δ~ID~ lies wholly outside ±0.010, and not materially
different if the 90% interval lies wholly within ±0.005, provided the observed
records and the no-validation method outputs are identical across the pair. An
estimator's bias is material if its 95% interval lies wholly outside ±0.010 and
equivalent if its 90% interval lies wholly within ±0.005.

# Results

## Structural contrast

The observed records and every no-validation method's output were identical
across the pair. Δ~ID~ was `r f(dec$delta_id, 5)`, with Monte Carlo standard
error `r formatC(dec$mcse, format = "e", digits = 1)` and a 90% interval of
`r f(dec$ci90_lo, 6)` to `r f(dec$ci90_hi, 6)`, wholly inside ±0.005. The
registered classification is not materially different.

Latent eligibility did change the population. In the decisive pair, the true
prevalence of eligibility was `r f(pe[["lambda_0"]])` at λ = 0,
`r f(pe[["lambda_log2"]])` at log 2 and `r f(pe[["lambda_log4"]])` at log 4.
The effect among the eligible hardly moved because the share of comorbid
patients among them changed by less than one percentage point: with effects of
−0.04 and −0.12, a Δ~ID~ of −0.0007 corresponds to a shift of about 0.008 in
that share.

## Estimators

Every estimator's bias for its declared risk difference was classed equivalent
in all 48 scenarios (@tbl-methods). The largest absolute bias at M = 20 or
without imputation was `r f(max(abs(full$bias)), 4)`, and coverage ran from
`r f(min(full$coverage))` to `r f(max(full$coverage))`. The complete-case
analysis estimates the effect among recorded eligibles, which differed from the
effect among all eligibles by at most `r f(max(abs(td$target_discrepancy)), 4)`.

```{r}
#| label: tbl-methods
#| tbl-cap: "Risk-difference performance across the 48 scenarios: largest absolute bias and the range of coverage of the nominal 95% interval. The M = 50 imputation check ran on 80 replicates per scenario and is not shown."
t <- do.call(rbind, lapply(split(full, full$mv), function(d) data.frame(
  method = d$mv[1], information = d$information_set[1],
  max_abs_bias = f(max(abs(d$bias)), 4),
  coverage = paste(f(min(d$coverage)), "to", f(max(d$coverage))))))
knitr::kable(t[order(t$information, t$method), ], row.names = FALSE)
```

Prevalence tells the other half (@tbl-prevalence). Without validation, the
missing-at-random estimators and the sensitivity analysis at λ\* = 0 recovered
prevalence at λ = 0 but underestimated it by up to
`r f(max(abs(mar$bias)))` at λ = log 4. The sensitivity analysis recovered it
only when its assumed λ\* equaled the true λ. The three validation-based
estimators were within `r f(max(abs(val$bias)), 4)` at every λ.

```{r}
#| label: tbl-prevalence
#| tbl-cap: "Bias in the prevalence of eligibility by the true λ, in the decisive setting (heterogeneous effect, m = 0.50, κ = 1.10, expected validation sample 100). The complete-case row is the recorded yield, which is not a prevalence estimator."
w <- tapply(pair$bias, list(pair$mv, pair$lambda_key), mean)
w <- w[!grepl("_m50", rownames(w)), c("lambda_0", "lambda_log2", "lambda_log4")]
colnames(w) <- c("λ = 0", "λ = log 2", "λ = log 4")
knitr::kable(round(w, 4))
```

# Discussion

The structural claim of ELG-01 holds: records alone cannot tell which of two
eligibility laws produced them. Whether that matters depends on the quantity.
The number of eligible people is not identified, and estimators that assume
eligibility is missing at random miss it by as much as the hidden shift moves
it. The risk difference among the eligible differs by less than 0.001 between
laws that produce the same records, because the effect differs across patients only through comorbidity and hidden
eligibility barely changes comorbidity's share. For the effect to change by
0.010 with these effect sizes, the share would have to move by about 12
percentage points.

A validation sample with known inclusion probabilities repairs both quantities.
Without one, an emulation can report the effect among the eligible with some
confidence and the size of the eligible population with none, unless the
missing measurement is itself strongly tied to who benefits.

## What this study does not establish

The result is conditional on one baseline threshold criterion, a binary
60-month outcome, correctly specified models and a single effect modifier. It
cannot estimate a real-world λ or establish empirical materiality. It does not
cover external, nonprobability or unknown-probability validation, validation
measured with error, time-varying eligibility, treatment switching, censoring
or unmeasured treatment confounding. A hidden eligibility mechanism tied more
strongly to the effect modifier than this one could change the effect
materially.

## Review

The design was reviewed and revised before any code was written. An earlier
version of the analysis called the bias-eliminated coverage function with its
arguments in the wrong order; that was fixed across all studies before this
run. The results have not yet been reviewed.

# Data and code availability {.unnumbered}

Code, protocol and result files: <https://github.com/choxos/TTE-open-problems/tree/main/studies/ELG-01-latent-eligibility-validation>.

# References {.unnumbered}
